```{r}
library(dplyr)
library(ggplot2)
```

We will be practicing with `dplyr` functions using the `starwars` dataset:

```{r}
starwars
head(starwars)
?starwars
```

## Subset Rows with `Filter`

`filter()` allows us to subset the rows of a dataset based on the values

```{r}
## Filter for rows with human Star Wars characters
filter(starwars, species == "Human")

## Save to a new object
humans <- filter(starwars, species == "Human")
humans
filter(humans, sex == "male")

## filter multiple columns at once
filter(starwars, species == "Human", sex == "male")
```

\
We can use different comparison ops to filter:

-   ! is "not":

-   == is equal

-   & is and

-   \| is or

-   \< is less than

-   \> is more than

-   is.na() returns TRUE if value is NA

```{r}
## return characters with a missing mass
filter(starwars, is.na(mass))
```

```{r}
## How to get all characters that are NOT human?

filter(starwars, species != "Human")
```

variable_name %in% vector_of_values returns TRUE if the value of variable_name is equal to one of the values in vector_of_values

```{r}

skin_color <- "grey"
skin_color %in% c("yellow", "fair", "tan") #returns false because grey is not in vector of colors

## How would we filter for characters with brown or blond hair?
filter(starwars, hair_color %in% c("blond","brown"))
```

grepl("characters_of_interest", variable_name) returns TRUE if the value of variable_name contains the characters_of_interest.

```{r}
filter(starwars, grepl("blond", hair_color))

## How would we get all characters with blue in their eyes

filter(starwars, grepl("blue", eye_color))

filter(starwars, eye_color == "blue")

?grepl
```

You can use `<`, \>, \<=, \>= to compare numeric variables or categorical variables

-   nominal categorical variables(character vectors) are ranked alphabetically

-   ordinal categorical variables(factors) are ranked based on the built in order

```{r}

## Star wars characters with a name starting with letters before "D" 

filter(starwars, name < "D")

str(diamonds)

filter(diamonds, cut > "Good") #ranking is based on levels of the ordered factor cut

"D" < "d" #false
"d" < "D" #true
```

between(x, left, right) is a shorthand for x \>= left & x \<= right

```{r}
between(5,1,7) #true because 5 is less than or equal to 7 and greater than or equal to 1

filter(starwars, between(mass, 50, 100))
```

```{r}
filter(starwars, is.na(hair_color))
filter(starwars, species != "droids", height < 100)
filter(starwars, between(birth_year, 100, 200))
```

## Order rows with `arrange`

`arrange()` reorders rows based on the values of column(s)

```{r}
arrange(starwars, name)
"a" < "B"
# order of character values aAbBcCdD

#order by species and birth year
arrange(starwars, species, birth_year)

#use desc() to get the birth years in descending order
arrange(starwars, species, desc(birth_year))
```

Practice

```{r}
arrange(starwars, desc(height))
arrange(starwars, height, species)
arrange(starwars, hair_color, eye_color, name)
```

##  Subset columns with `select()`

```{r}
select(starwars, hair_color, skin_color, eye_color)
character_colors <- select(starwars, hair_color, skin_color, eye_color)
character_colors
```

We can also select multiple columns in a row

```{r}
select(starwars, name : eye_color) # all columns between name and eye_color(inclusive)
```

We can also remove columns by name using `-`

```{r}
starwars
select(starwars, -films, -vehicles, -starships)
select(starwars, -c(films, vehicles, starships))
select(starwars, -c(films:starships))
```

Helpful functions

`everything()` is a shortcut for everything else

```{r}
select(starwars, "species", everything()) #put specieis first, then all the other columns

## alphabetize columns
starwars_names <- sort(names(starwars))
starwars_names
select(starwars, starwars_names)
```

stars_with("abc") matches column names that start with abc

ends_with("abc") matches column names that end with abc

contains("abc") matches column names that end with "abc"

```{r}
select(starwars, starts_with("h"))

select(starwars, contains("Or"))
```

Practice!

```{r}
select(starwars, "species", "homeworld")

#2 
select(starwars, contains("_"))
select(starwars, contains("color"), birth_year)
```

## Add new variables with `mutate`

```{r}
starwars_small <- select(starwars, height, mass, species)
starwars_small

mutate(starwars_small, height_m = height/100)
```

We can also create multiple variables at once and we can refer to variables we just created

```{r}
mutate(starwars_small, height_m = height/100, bmi = mass/height_m^2)

##only keep new variables
transmute(starwars_small, height_m = height/100, bmi = mass/height_m ^2)

```

Summary statistics give one value for the entire column: sum(), mean()

```{r}
mutate(starwars_small, sum = sum(mass, na.rm = T))

mutate(starwars_small, row_addition = height + mass, summary_sum = sum(height, mass, na.rm = T))
```

Group by a variable using the `.by` parameter

```{r}
mutate(starwars_small, species_mass = mean(mass, na.rm = T), .by = species)
```

```{r}
mutate(starwars_small, mass_lbs = mass*2.2)
```

Change the type of a variable

```{r}
mutate(starwars_small, mass = as.character(mass))
```

## Summarize data with `summarise`

```{r}
summarise(starwars, mean_mass = mean(mass, na.rm = T))

# Group by species

summarise(starwars, mean_mass = mean(mass, na.rm = T), .by = species)
```

The only columns kept are the columns you group by and the columns you create

`n()` is a helpful function that returns the number of rows in a group

```{r}
summarise(starwars, mean_mass = mean(mass, na.rm = T), count = n(), .by = species)

# group by multiple groups
summarise(starwars, mean_mass = mean(mass, na.rm = T), count = n(), .by = c(species, sex))
```
